Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-05-14 Chapter_1/04_Manuscript/Manuscript.tex
worked on manuscript

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Step by Step: Using Step Selection Analysis to Simulate Dispersal and
Assess Landscape Connectivity}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification of dispersal corridors has become
a task of extraordinary importance in ecological studies.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost methods and circuit theory to investigate landscape
connectivity. Despite their usefulness for various ecological applications, both
methods make several restricting assumptions that limit their suitability in
reality. To address their shortcomings, simulations from individual-based
movement models have been proposed and applied. To overcome their shortcomings,
simulations from individual-based movement models have been proposed. Yet, due
to the infinite number of possibilities of coming up with an individual-based
model, a unified and objective framework is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and landscape connectivity. In particular, the rich
suite of resource selection functions, including point-, step-, and
path-selection functions have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which individual animal movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution, a similar approach may be useful for investigating
\textit{transient} movement behavior and thereby study landscape connectivity by
means of simulated dispersal events.

Here, we showcase the use of integrated step selection analysis to simulate
dispersal of the highly endangered African wild dog across the world's largest
transboundary conservation area, the Kavango-Zambezi Transfrontier Conservation
Area (KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions departing from northern Botswana. We combine the movement data with
relevant spatial covariate layers and used integrated step selection functions
to parametrize a fully mechanistic movement model rendering wild dog dispersal.
Based on this model, we simulate 80'000 dispersers, originating from protected
areas and moving across the extent of the KAZA-TFCA. We then generate a heatmap
and use network theory to reveal dispersal hotspots and crucial bottlenecks
across the study area. Finally, we discuss the benefits and pitfalls of such
dispersal simualtions and highlight potential improvements to be made in the
future.
\end{abstract}

@@

\section{Introduction (90\%)}

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals from their natal location to
the site of first reproduction \cite{Howard.1960}. It is a vital process
governing the dynamics wild animal populations that are distributed in space
\citep{Hanski.1998, Clobert.2012} and may strongly affect population dynamics at
different spatial and social scales \citep{Hanski.1999a, Clobert.2012}.
Dispersal allows species to avoid inbreeding and maintain genetic diversity
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013}, to
rescue small and non-viable populations \citep{Brown.1977}, and to promote the
colonization of unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}.
However, successful dispersal requires a sufficient degree of functional
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become an important task in
conservation science \citep{Doerr.2011, Rudnick.2012}. To achieve this, reliable
information on movement behavior during dispersal and knowledge about factors
that limit dispersal and connectivity is paramount \citep{Baguette.2013,
Vasudev.2015}.

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modelling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
path analysis (LCP analysis; \citealp{Adriaensen.2003}) and circuit theory (CT
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape. Despite their intuitive nature and ease of use,
both methods make rigorous assumptions about animal movement that may or may not
be fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths \& Circuit Theory (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range, a preconceived end-point in mind, and choose a cost-minimizing
route accordingly. These assumptions may be reasonable for migrating animals,
yet they are unlikely to be true for dispersers that move over long distances
into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Finally,
LCPs are only one pixel wide, meaning that their absolute size depends on the
resolution of chosen covariate layers \citep{Diniz.2020}. Although some of these
issues can be alleviated using less stringent versions of the LCP algorithm
(e.g. least-cost \textit{corridors} \citep{Pinto.2009}, \textit{thresholded}
least-cost paths \citep{Landguth.2012}, and \textit{randomized} least-cost paths
\citep{Panzacchi.2016, VanMoorter.2021}), a certain degree of arbitrariness in
the assumptions remains. CT entails similar restrictions. Because CT only
considers movements from the source cell to its 4 or 8 adjacent cells, it
implicitly posits a perceptual range of a single pixel. It is also built on the
assumption of a complete random walk \citep{Diniz.2020}, implying that
directional biases cannot be rendered, albeit being very common in dispersal
movements \citep{Cozzi.2020, Hofmann.2021}. Ultimately, both LCP analysis and CT
omit the temporal dimension of dispersal. In result, statements about the
expected duration required to traverse a certain corridor are impossible.
Because movement is not modelled explicitely, neither of the methods can render
possible interactions between movement behavior and landscape characteristics
such that connectivity mainly arises as a result of the landscape structure.
This is referred to as structural connectivity and stands in contrast to
functional connectivity, which also renders the behavioral response of the
animal with respect to prevailing habitat conditions \citep{Tischendorf.2000}.
Even though functional connectivity is more difficult to estimate, a functional
view is the ultimate goal in conservation science because it has direct
consequences for gene flow \citep{Baguette.2013}.

@@

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient
landscape connectivity is the endangered African wild dog \textit{Lycon pictus}.
While once present across entire sub-Saharan Africa, wild dogs have disappeared
from a vast majority of their historic range due to persecution by humans,
habitat destruction, and deadly diseases. As of today, only 6'000 free-ranging
individuals remain in small and spatially scattered subpopulations
\citep{Woodroffe.2012}. Within those subpopulations, wild dogs form cohesive
packs comprising 8 to 12 adults and their offspring \cite{McNutt.1995}. After
reaching sexual maturity, male and female offspring form same-sex coalitions and
disperse from their natal pack \citep{McNutt.1996, Behr.2020}. New packs are
formed when dispersing coalitions join unrelated opposite-sex dispersing
coalitions \citep{McNutt.1996}. Dispersing wild dogs can cover several hundred
kilometers across a variety of landscapes \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the few strongholds for this
species lies near the Moremi Game Reserve in northern Botswana, which is part of
the world's largest transboundary protected area, namely the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). This area has originally been
intended to facilitate migration of elephants, but is expected to benefit a
multitude of other species \citep{Elliot.2014, Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost corridor approach
\citep{Hofmann.2021}. For this, we fitted a basic habitat selection model based
on which we predicted landscape permeability. We now expand on this knowledge
and use ISSF to develop a more detailed movement model of dispersing wild dogs.
We then use this model to simulate dispersers moving across the KAZA-TFCA. Based
on simulations, we compute heatmaps and identify potential dispersal hotspots.
We also showcase how network metrics relevant to landscape connectivity can be
computed. Our results show that a simulation-based approach yields several major
benefits over traditional connectivity modeling techniques. Most importantly,
simulations provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addtion, by generating
proper dispersal trajectories, network theory can be applied to calculate
network metrics that are pertinent to connectivity analysis. Finally, we put
forward additional opportunities using simulations that go beyond the scope of
this paper.

\section{Methods}
\subsection{Study Area (90\%)}
The study area was centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) and was represented by a rectangular bounding box that
stretched over 1.3 Mio. km\textsuperscript{2}, ecompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, covering a total of 520'000 km\textsuperscript{2}. Its landscape
varies regionally and ranges from savanna, to grassland and from dry to moist
woodland habitats. A dominant hydrogeographical feature in its center is the
flood-pulsing Okavango Delta. The wet season within the KAZA-TFCA lasts from
November to March and is out of phase with the main flooding of the Okavango
Delta which peaks between July and August \citep{McNutt.1996, Wolski.2017}.
While large portions within the KAZA-TFCA are designated national parks or other
protected areas, substantial human influence remains due to roads, agricultural
sites and settlements and villages.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{(a) The study area was confined by a bounding box encompassing the
    entire KAZA-TFCA and spanned parts of Angola, Namibia, Botswana, Zimbabwe,
    and Zambia. (b) The KAZA-TFCA represents the world's largest terrestrial
    conservation area and covers a total of 520'000 km\textsuperscript{2}. Its
    purpose is to re-establish connectivity between already-existing national
    parks (dark green) and other protected areas (light green). The dispersal
    data used in this study was collected on a free-ranging African wild dog
    population inhabiting the Moremi National Park in northern Botswana.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data (90\%)}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana. We selected potential dispersers based on age, pack size,
number of same‐sex siblings within the pack, and presence of unrelated
opposite-sex individuals in the pack \citep{McNutt.1996, Behr.2020}. We
immobilized selected individuals using a cocktail of Ketamine/Xylazine/Atropine
\citep{Osofsky.1996, Cozzi.2020} that was injected by dart, fired from a
CO\textsubscript{2}-pressurized gun (\textit{DAN-Inject, Denmark}). Immobilized
individuals were fitted with GPS/Satellite radio collars (\textit{Vertex Lite;
Vectronic Aerospace GmbH, Berlin}) that guaranteed automated drop-off through a
decomposable piece of cotton. Handling and collaring of all individuals was
supervised by a Botswana-registered wildlife veterinarian. All individuals
rejoined their pack within one hour after immobilization. 16 collared
individuals eventually dispersed, each in a separate same-sex dispersal
coalition (7 female and 9 male coalitions).

Because behavior during dispersal is more pertinent for assessing landscape
connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data that was
collected during residency and only retained GPS data recorded during dispersal.
During dispersal, collars were programmed to record a GPS fix every 4 hours.
Collected relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. In some instances, exact dispersal
dates were known from field observations. Otherwise, we distinguished between
residency and dispersal using the net-squared displacement metric. This metric
measures the squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. As such, dispersal was deemed to have started
when an individual left its natal home range and ended once individuals became
sedentary again. As previous research revealed similar behavior of females and
males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not
distinguish between sexes. After collection, we converted collected GPS
coordinates (n = 4'169) to steps, where each step represented the straight-line
distance traveled by and individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule.

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates including water-cover, distance to water, woodland-cover, and
shrub/grassland-cover. Because water cover greatly changes within and between
years in the Okavango Delta, we applied a remote sensing algorithm and generated
frequently updated water cover layers and corresponding distance to water layers
(see \citealp{Wolski.2017} and \citealp{Hofmann.2021}). Resulting layers thus
temporally aligned with our dispersal events. We furthermore computed a proxy
for human influence, depicting anthropogenic pressures stemming from
human-density, agricultural sites, and roads. All spatial layers were coarsened
or interpolated to a target resolution of 250 m by 250 m. Further details on the
sources and preparation of each habitat covariate are given in
\cite{Hofmann.2021}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}). Because wild dogs follow a diurnal activity pattern, we also
coded a binary variable (\textsf{LowActivity}) indicating whether a step was
realized during periods of low wild dog activity (17:00 to 07:00 local time) or
high wild dog activity (09:00 to 17:00 local time). Handling and manipulation of
all data, as well as all models and simulations were implemented with the
statistical software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper
functions were written in {\tt C++} and imported into {\tt R} using the {\tt
Rcpp} package \citep{Eddelbuettel.2011, Eddelbuettel.2013}

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct iSSF analysis, we paired each observed step
with 24 random steps. An observed and its 24 random steps thus formed a stratum
and received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to observed steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)}. To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. Since correlation among covariates was low (\(|r| > 0.6\);
\citealp{Latham.2011}), we retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step. The selection score \(w(x)\) of each step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's relative selection
strengths (i.e. preferences) towards these covariates (\(\beta_1, \beta_2, ...,
\beta_n\)):

@@

To estimate the preferences of interest, we ran conditional logistic regression
models in the r-package {\tt glmmTMB}. To handle multiple individuals, we
applied the mixed effects method developed by \citep{Muff.2020}, which allows to
fit random slopes in addition to random intercepts. Thus, we treated animal IDs
as random effect and modeled random slopes for each covariate. We fixed the
random intercept variance at an arbitrary high value of 10\textsuperscript{6} to
make use of the  ``poission''-trick \citep{Muff.2020}.

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Step by Step: Using Step Selection Functions to Simulate Dispersal and
Assess Landscape Connectivity}

% \title{Step by Step: The Utility of Dispersal Simulations to Assess Landscape
% Connectivity}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification of dispersal corridors has become
a task of extraordinary importance for conservation authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. To address these shortcomings, dispersal
simulations from individual-based movement models have been proposed and
applied. Yet, due to the almost infinite amount of non-trivial decisions a
modeler faces when parametrizing such models, a unified and objective framework
is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful for
investigating \textit{transient} movement behavior and study landscape
connectivity.

Here, we showcase the use of integrated step selection functions to simulate
dispersal of the endangered African wild dog across the world's largest
transboundary conservation area, the Kavango-Zambezi Transfrontier Conservation
Area (KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions and parametrize a fully mechanistic
movement model rendering wild dog dispersal. Based on this model, we simulate
80'000 dispersers, originating from protected areas and moving across the extent
of the KAZA-TFCA. We then generate a heatmap indicating regions frequently
visited by dispersers, use network theory to reveal dispersal hotspots and
crucial bottlenecks across the study area. Finally, we discuss the benefits and
pitfalls of such dispersal simualtions and highlight potential improvements to
be made in the future.

\end{abstract}

@@

\section{Introduction}

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the dynamics wild animal populations that are distributed in
space \citep{Hanski.1998, Clobert.2012} and may strongly affect population
dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to avoid inbreeding and maintain genetic
diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, to rescue small and non-viable populations \citep{Brown.1977},
and to promote the colonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
functional connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become an
important task in conservation science \citep{Doerr.2011, Rudnick.2012}. In
order to pinpoint relevant dispersal hotspots, reliable information on movement
behavior during dispersal and knowledge about factors that limit dispersal and
connectivity is paramount \citep{Baguette.2013, Vasudev.2015}.

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modelling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. In this regard, two of the most prominent connectivity
models are least-cost path analysis (LCP analysis; \citealp{Adriaensen.2003})
and circuit theory (CT \citealp{McRae.2006, McRae.2008}), both graph-based
methods that estimate conductance of the landscape. Despite their intuitive
nature and ease of use, both methods make rigorous assumptions about animal
movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths \& Circuit Theory (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range, a preconceived end-point in mind, and choose a cost-minimizing
route accordingly. These assumptions may be reasonable for migrating animals,
yet they are unlikely to hold for dispersers, which typically move over long
distances into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}.
Finally, LCPs are only one pixel wide, meaning that their absolute size depends
on the resolution of chosen covariate layers \citep{Diniz.2020}. Although some
of these deficiencies can be addressed using less stringent versions of the LCP
algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness in the assumptions remains. CT entails similarly
unreasonable restrictions. Because CT only considers movements from the source
cell to its 4 or 8 adjacent cells, it implicitly posits a perceptual range of a
single pixel. As such, the perceptual range cannot be defined based on
biological observations, but is imposed by the resolution of the reference grid.
CT is also built on the assumption of a complete random walk \citep{Diniz.2020},
implying that directional biases cannot be rendered, albeit being very common in
dispersal movements \citep{Cozzi.2020, Hofmann.2021}. Ultimately, neither LCP
analysis nor CT are capable of rendering the temporal dimension of dispersal.
Statements about the expected duration required to traverse a certain corridor
are therefore impossible, albeit being important factors. Likewise, because
movement is not modelled explicitely, neither of the methods allows to render
interactions between movement and habitat preferences of the focal species. This
implies that connectivity mainly arises as a result of the landscape structure
and is referred to as structural connectivity. Structural connectivity stands in
contrast to functional connectivity, which also renders the behavioral response
of the animal with respect to prevailing habitat conditions
\citep{Tischendorf.2000}. Even though functional connectivity is more difficult
to estimate, a functional view is the ultimate goal in conservation science
because it has direct consequences for gene flow \citep{Baguette.2013}.

@@

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012,
Woodroffe.2020}. As of today, only 6'000 free-ranging individuals remain in
small and spatially scattered subpopulations \citep{Woodroffe.2012}. Within
those subpopulations, wild dogs form cohesive packs comprising 8 to 12 adults
and their offspring \cite{McNutt.1995}. After reaching sexual maturity, male and
female offspring form same-sex coalitions and disperse from their natal pack
\citep{McNutt.1996, Behr.2020}. New packs are formed when dispersing coalitions
join unrelated opposite-sex dispersing coalitions \citep{McNutt.1996}.
Dispersing wild dogs can cover several hundred kilometers across a variety of
landscapes \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}.
One of the few strongholds for this species lies near the Moremi Game Reserve in
northern Botswana, which is part of the world's largest transboundary protected
area, namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).
This area has originally been intended to facilitate migration of elephants, but
is expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost corridor approach
\citep{Hofmann.2021}. For this, we fitted a basic habitat selection model based
on which we predicted landscape permeability. We now expand on this knowledge
and use ISSF to develop a more detailed movement model of dispersing wild dogs
(\Cref{GraphicalAbstract}). We then use this model to simulate dispersers moving
across the KAZA-TFCA. Based on simulations, we compute heatmaps and identify
potential dispersal hotspots. We also showcase how network metrics relevant to
landscape connectivity can be computed. Our results show that a simulation-based
approach yields several major benefits over traditional connectivity modeling
techniques. Most importantly, simulations provide a more generic view on how
connectivity emerges and to which degree connectivity depends on the dispersal
duration. In addtion, by generating proper dispersal trajectories, network
theory can be applied to calculate network metrics that are pertinent to
connectivity analysis.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis. First, GPS
    data and habitat covariates have to be collected. The combined data is then
    analyzed in an integrated step selection model, which enables the
    parametrization of the focal species' habitat and movement kernels. The
    parametrized model is then treated as an individual-based movement model and
    used to simulate dispersal trajectories. Ultimately, simulated trajectories
    serve to produce a set of maps that are pertinent to landscape connectivity.
    This includes a heatmap, indicating the relative traversal frequency across
    the study area, a betweenness map, highlighting movement corridors and
    bottlenecks, and, finally, an inter-patch connectivity map, where the
    frequency of connections and their average duration can be depicted. Photos:
    Whom to cite? Vectronics or Photographers?}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and ecompassing the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts a diverse landscape, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    encompassing parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green). The dispersal data used in
    this study was collected on a free-ranging African wild dog population
    inhabiting the Moremi National Park in northern Botswana.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data (90\%)}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana \citep{Cozzi.2020, Hofmann.2021}. We selected potential
dispersers based on age, pack size, number of same‐sex siblings within the pack,
and presence of unrelated opposite-sex individuals in the pack
\citep{McNutt.1996, Behr.2020}. We immobilized selected individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020} that was
injected by dart, fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Immobilized individuals were fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}) that guaranteed automated drop-off through a decomposable piece of
cotton. Handling and collaring of all individuals was supervised by a
Botswana-registered wildlife veterinarian and all individuals quickly rejoined
their pack after immobilization.

16 collared individuals eventually dispersed, each in a separate same-sex
dispersal coalition (7 female and 9 male coalitions). During dispersal, collars
were programmed to record a GPS fix every 4 hours. Collected relocations were
regularly transmitted over the Iridium satellite system, which allowed remote
tracking of individuals, even if they left the main study area and crossed
international borders. Because behavior during dispersal is more pertinent for
assessing landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded
all data that was collected during residency and only retained GPS data recorded
during dispersal. In some instances, exact dispersal dates were known from field
observations. Where this was not the case, determined dispersal phases using the
net-squared displacement metric. Net squared displacement measures the squared
Euclidean distance of a GPS relocation to a reference point \citep{Borger.2012},
which in our case was set to the center of each individual's natal home range.
As such, dispersal was deemed to have started when an individual left its natal
home range and ended once individuals became sedentary again. As previous
research revealed similar behavior of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by and individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not successfully collected
on the 4-hourly schedule (\( \pm \) 15 minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting water layers thus temporally aligned with our
dispersal events. We furthermore computed a proxy for human influence, rendering
anthropogenic pressures stemming from human-density, agricultural sites, and
roads. All spatial layers were coarsened or interpolated to a target resolution
of 250 m by 250 m. Further details on the sources and preparation of each
habitat covariate are given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern, we also coded a binary variable
(\textsf{LowActivity}) indicating whether a step was realized during periods of
low wild dog activity (17:00 to 09:00 local time) or high wild dog activity
(09:00 to 17:00 local time). Handling and manipulation of all data, as well as
all models and simulations were implemented with the statistical software {\tt
R}, version 3.6.6 \citep{R.2019}. Several helper functions were written in {\tt
C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed and its 24 random steps formed a stratum and
received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to realized steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox} package \citep{Hunziker.2017}. We also calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
observed and random step. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one. Since
correlation among covariates was low (\(|r| > 0.6\); \citealp{Latham.2011}), we
retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step \citep{Fortin.2005}. The selection score \(w(x)\) of each step depended on
its associated covariates (\(x_1, x_2, ..., x_n\)) and on the animal's
preferences (i.e. relative selection strengths; \citealp{Avgar.2017}) towards
these covariates (\(\beta_1, \beta_2, ..., \beta_n\)):

@@

To estimate the preferences of interest, we ran conditional logistic regression
analysis in the r-package {\tt glmmTMB}. To handle multiple individuals, we
applied the mixed effects technnique developed by \citep{Muff.2020}, which
allows to effectively model random slopes. Thus, we treated animal IDs as random
effect and modeled random slopes for each covariate. We fixed the random
intercept variance at an arbitrary high value of 10\textsuperscript{6} to make
use of the  ``poission''-trick \citep{Muff.2020}.
